---
title: "Pull the (character) strings with stringi 0.5-2"
author: "Marek Gagolewski"
date: "2015-06-23"
output:
  html_document:
    theme: united
    highlight: tango
---

A reliable string processing toolkit is a must-have for any data scientist.

A new release of the [`stringi`](https://github.com/Rexamine/stringi/) package is
available on [CRAN](http://cran.r-project.org/web/packages/stringi/index.html)
(please wait a few days for Windows and OS X binary builds). 
As for now, about 850 CRAN packages depend (either directly or recursively)
on `stringi`. And quite recently, the package got listed
among the [top downloaded R extensions](http://www.r-pkg.org/downloaded).


```{r}
# install.packages("stringi") or update.packages()
library("stringi")
```

```{r,cache=TRUE}
stri_info(TRUE)
apkg <- available.packages(contriburl="http://cran.rstudio.com/src/contrib")
length(tools::dependsOnPkgs('stringi', installed=apkg, recursive=TRUE))
```

Refer to the [`INSTALL`](http://cran.r-project.org/web/packages/stringi/INSTALL)
file for more details if you compile stringi from sources
(Linux users mostly).

Here's a list of [changes](http://cran.r-project.org/web/packages/stringi/NEWS)
in version 0.5-2.
There are many major (like date&time processing) and minor new features,
enhancements, as well as bugfixes.
In the current release we also
focused on bringing [`stringr`](https://github.com/hadley/stringr)
package's users even better string processing experience,
as since the 1.0.0 release it is now 
[powered by `stringi`](http://blog.rstudio.org/2015/05/05/stringr-1-0-0/).

* [BACKWARD INCOMPATIBILITY] The second argument to `stri_pad_*()` has
been renamed `width`.

* [GENERAL] #69: `stringi` is now bundled with [ICU4C 55.1](http://site.icu-project.org/download/55).

* [NEW FUNCTIONS] #137: date-time formatting/parsing (note that this is draft API
and it may change in future `stringi` releases; any comments are welcome):
     * `stri_timezone_list()` - lists all known time zone identifiers
    ```{r}
    sample(stri_timezone_list(), 10)
    ```
     * `stri_timezone_set()`, `stri_timezone_get()` - manage current default time zone
     * `stri_timezone_info()` - basic information on a given time zone
    ```{r}
    str(stri_timezone_info('Europe/Warsaw'))
    stri_timezone_info('Europe/Warsaw', locale='de_DE')$Name
    ```
     * `stri_datetime_symbols()` - localizable date-time formatting data
    ```{r}
    stri_datetime_symbols()
    stri_datetime_symbols("th_TH_TRADITIONAL")$Month
    stri_datetime_symbols("he_IL@calendar=hebrew")$Month
    ```
     * `stri_datetime_now()` - return current date-time
     * `stri_datetime_fstr()` - convert a `strptime`-like format string
            to an ICU date/time format string
     * `stri_datetime_format()` - convert date/time to string
    ```{r}
        stri_datetime_format(stri_datetime_now(), "datetime_relative_medium")
    ```
     * `stri_datetime_parse()` - convert string to date/time object
    ```{r}
    stri_datetime_parse(c("2015-02-28", "2015-02-29"), "yyyy-MM-dd")
    stri_datetime_parse(c("2015-02-28", "2015-02-29"), stri_datetime_fstr("%Y-%m-%d"))
    stri_datetime_parse(c("2015-02-28", "2015-02-29"), "yyyy-MM-dd", lenient=TRUE)
    stri_datetime_parse("19 lipca 2015", "date_long", locale="pl_PL")
    ```
     
     * `stri_datetime_create()` - construct date-time objects
            from numeric representations
    ```{r}
    stri_datetime_create(2015, 12, 31, 23, 59, 59.999)
    stri_datetime_create(5775, 8, 1, locale="@calendar=hebrew") # 1 Nisan 5775 -> 2015-03-21
    stri_datetime_create(2015, 02, 29)
    stri_datetime_create(2015, 02, 29, lenient=TRUE)
    ```
     
     * `stri_datetime_fields()` - get values for date-time fields
    ```{r}
    stri_datetime_fields(stri_datetime_now())
   stri_datetime_fields(stri_datetime_now(), locale="@calendar=hebrew")
   stri_datetime_symbols(locale="@calendar=hebrew")$Month[
      stri_datetime_fields(stri_datetime_now(), locale="@calendar=hebrew")$Month
   ]
    ```
     * `stri_datetime_add()` - add specific number of date-time units
            to a date-time object
    ```{r}
    x <- stri_datetime_create(2015, 12, 31, 23, 59, 59.999)
    stri_datetime_add(x, units="months") <- 2
    print(x)
    stri_datetime_add(x, -2, units="months")
    ```

* [NEW FUNCTIONS] `stri_extract_*_boundaries()` extract text between text
boundaries.

* [NEW FUNCTION] #46: `stri_trans_char()` is a `stringi`-flavoured
`chartr()` equivalent.

```{r}
stri_trans_char("id.123", ".", "_")
stri_trans_char("babaab", "ab", "01")
```

* [NEW FUNCTION] #8: `stri_width()` approximates the *width* of a string
in a more Unicodish fashion than `nchar(..., "width")`

```{r}
stri_width(LETTERS[1:5])
nchar(stri_trans_nfkd("\u0105"), "width") # provides incorrect information
stri_width(stri_trans_nfkd("\u0105")) # A and ogonek (width = 1)
stri_width( # Full-width equivalents of ASCII characters:
   stri_enc_fromutf32(as.list(c(0x3000, 0xFF01:0xFF5E)))
)
```

* [NEW FEATURE] #149: `stri_pad()` and `stri_wrap()` now by default bases on 
code point widths instead of the number of code points. Moreover, the default
behavior of `stri_wrap()` is now such that it does not get rid
of non-breaking, zero width, etc. spaces

```{r}
x <- stri_flatten(c(
   stri_dup(LETTERS, 2),
   stri_enc_fromutf32(as.list(0xFF21:0xFF3a))
), collapse=' ')
# Note that your web browser may have problems with properly aligning
# this (try it in RStudio)
cat(stri_wrap(x, 11), sep='\n')
```

* [NEW FEATURE] #133: `stri_wrap()` silently allows for `width <= 0`
(for compatibility with `strwrap()`).

* [NEW FEATURE] #139: `stri_wrap()` gained a new argument: `whitespace_only`.

* [GENERAL] #144: Performance improvements in handling ASCII strings
(these affect `stri_sub()`, `stri_locate()` and other string index-based
 operations)

* [GENERAL] #143: Searching for short fixed patterns (`stri_*_fixed()`) now
relies on the current `libC`'s implementation of `strchr()` and `strstr()`.
This is very fast e.g. on `glibc` utilizing the `SSE2/3/4` instruction set.

```{r}
x <- stri_rand_strings(100, 10000, "[actg]")
microbenchmark::microbenchmark(
   stri_detect_fixed(x, "acgtgaa"),
   grepl("actggact", x),
   grepl("actggact", x, perl=TRUE),
   grepl("actggact", x, fixed=TRUE)
)
```

* [GENERAL] #141: a local copy of `icudt*.zip` may be used on package
install; see the [`INSTALL`](http://cran.r-project.org/web/packages/stringi/INSTALL) file for more information.

* [GENERAL] #165: the `./configure` option `--disable-icu-bundle`
forces the use of system ICU when building the package.

* [BUGFIX] locale specifiers are now normalized in a more intelligent way:
e.g. `@calendar=gregorian` expands to `DEFAULT_LOCALE@calendar=gregorian`.

* [BUGFIX] #134: `stri_extract_all_words()` did not accept `simplify=NA`.

* [BUGFIX] #132: incorrect behavior in `stri_locate_regex()` for matches
of zero lengths.

* [BUGFIX] stringr/#73: `stri_wrap()` returned `CHARSXP` instead of `STRSXP`
on empty string input with `simplify=FALSE` argument.

* [BUGFIX] #164: libicu-dev usage used to fail on Ubuntu.

* [BUGFIX] #135: C++11 is now used by default (see the `INSTALL` file,
however) to build stringi from sources. This is because ICU4C uses the
`long long` type which is not part of the C++98 standard.

* [BUGFIX] #154: Dates and other objects with a custom class attribute
were not coerced to the character type correctly.

* [BUGFIX] #168: Build now fails if `icudt` is not available.

* [BUGFIX] Force ICU `u_init()` call on stringi dynlib load.

* [BUGFIX] #157: many overfull hboxes in the package PDF manual has been
corrected.

 
Enjoy! Any comments and suggestions are [welcome](https://github.com/Rexamine/stringi/issues).
