---
title: "Faster, easier, and more reliable character string processing with stringi 0.3-1"
author: "Marek Gagolewski"
date: "06.11.2014"
output: html_document
---

A new release of the [`stringi`](http://stringi.rexamine.com) package is
available on [CRAN](http://cran.r-project.org/web/packages/stringi/index.html)
(please wait a few days for Windows and OS X binary builds).

```{r}
# install.packages("stringi") or update.packages()
library("stringi")
```

**[`stringi`](http://stringi.rexamine.com) is an R package providing** (but definitely not limiting to) equivalents
of nearly all the **character string processing functions** known
from base R. While developing the package we had
**high performance** and [**portability**](http://www.rexamine.com/2014/05/stringi-0-2-3-released/)
of its facilities in our minds.

We implemented each string processing function from scratch.
The internationalization and globalization
support, as well as many string processing facilities
(like regex searching) is guaranteed by the well-known
**IBM's [ICU4C](http://site.icu-project.org/) library**.

Here is a very general list of the **most important
features** available in the current version of [`stringi`](http://stringi.rexamine.com):

* string searching:
    * with ICU (Java-like) regular expressions,
    * ICU USearch-based locale-aware string searching
      (quite slow, but working properly e.g. for non-Unicode normalized strings),
    * very fast, locale-independent byte-wise pattern matching;
* joining and duplicating strings;
* extracting and replacing substrings;
* string trimming, padding, and text wrapping (e.g. with Knuth's dynamic word wrap algorithm);
* text transliteration;
* text collation (comparing, sorting);
* text boundary analysis (e.g. for extracting individual words);
* random string generation;
* Unicode normalization;
* character encoding conversion and detection;

and many more.

Here's a list of changes in version 0.3-1:

* **(IMPORTANT CHANGE)** [#87](https://github.com/Rexamine/stringi/issues/87): 
  `%>%` overlapped with the pipe operator from
  the [`magrittr`](http://cran.r-project.org/web/packages/magrittr/index.html)
  package; now each operator like `%>%` has been renamed `%s>%`.

* **(IMPORTANT CHANGE)** [#108](https://github.com/Rexamine/stringi/issues/108):
  Now the BreakIterator (for text boundary analysis)
  may be better controlled via `stri_opts_brkiter()` (see options `type`
  and `locale` which aim to replace now-removed `boundary` and `locale` parameters
  to `stri_locate_boundaries`, `stri_split_boundaries`, `stri_trans_totitle`,
  `stri_extract_words`, `stri_locate_words`).
  
    For example:

```{r}
test <- "The\u00a0above-mentioned    features are very useful. Warm thanks to their developers. 123 456 789"
stri_split_boundaries(test, stri_opts_brkiter(type="word", skip_word_none=TRUE, skip_word_number=TRUE)) # cf. stri_extract_words
stri_split_boundaries(test, stri_opts_brkiter(type="sentence")) # extract sentences
stri_split_boundaries(test, stri_opts_brkiter(type="character")) # extract characters
```

By the way, the last call also works correctly for strings not in the Unicode
[Normalization Form C](http://docs.rexamine.com/R-man/stringi/stri_trans_nf.html):

```{r}
stri_split_boundaries(stri_trans_nfkd("zażółć gęślą jaźń"), stri_opts_brkiter(type="character"))
```


* **(NEW FUNCTIONS)** [#109](https://github.com/Rexamine/stringi/issues/109):
  `stri_count_boundaries` and `stri_count_words`
  count the number of text boundaries in a string.
  
```{r}
stri_count_words("Have a nice day!")
```

* **(NEW FUNCTIONS)** [#41](https://github.com/Rexamine/stringi/issues/41):
  `stri_startswith_*` and `stri_endswith_*`
  determine whether a string starts or ends with a given pattern.
  
```{r}
stri_startswith_fixed(c("a1o", "a2g", "b3a", "a4e", "c5a"), "a")
```

* **(NEW FEATURE)** [#102](https://github.com/Rexamine/stringi/issues/102):
  `stri_replace_all_*` gained a `vectorize_all` parameter,
  which defaults to TRUE for backward compatibility.
  
```{r}
stri_replace_all_fixed("The quick brown fox jumped over the lazy dog.",
     c("quick", "brown", "fox"), c("slow",  "black", "bear"), vectorize_all=FALSE)

# Compare the results:
stri_replace_all_fixed("The quicker brown fox jumped over the lazy dog.",
     c("quick", "brown", "fox"), c("slow",  "black", "bear"), vectorize_all=FALSE)
stri_replace_all_regex("The quicker brown fox jumped over the lazy dog.",
     "\\b"%s+%c("quick", "brown", "fox")%s+%"\\b", c("slow",  "black", "bear"), vectorize_all=FALSE)
```

* **(NEW FUNCTIONS)** [#91](https://github.com/Rexamine/stringi/issues/91):
  `stri_subset_*`, a convenient and more efficient
  substitute for `str[stri_detect_*(str, ...)]`, added.
  
```{r}
stri_subset_regex(c("john@office.company.com", "steve1932@g00gl3.eu", "no email here"),
   "^[A-Za-z0-9._%+-]+@([A-Za-z0-9-]+\\.)+[A-Za-z]{2,4}$")
```

* **(NEW FEATURE)** [#100](https://github.com/Rexamine/stringi/issues/100):
  `stri_split_fixed`, `stri_split_charclass`,
  `stri_split_regex`, `stri_split_coll` gained a `tokens_only` parameter,
  which defaults to `FALSE` for backward compatibility.

```{r}
stri_split_fixed(c("ab_c", "d_ef_g", "h", ""), "_", n_max=1, tokens_only=TRUE, omit_empty=TRUE)
stri_split_fixed(c("ab_c", "d_ef_g", "h", ""), "_", n_max=2, tokens_only=TRUE, omit_empty=TRUE)
stri_split_fixed(c("ab_c", "d_ef_g", "h", ""), "_", n_max=3, tokens_only=TRUE, omit_empty=TRUE)
```

* **(NEW FUNCTION)** [#105](https://github.com/Rexamine/stringi/issues/105):
  `stri_list2matrix` converts lists of atomic vectors
  to character matrices, useful in connection with `stri_split`
  and `stri_extract`.
  
```{r}
stri_list2matrix(stri_split_fixed(c("ab_c", "d_ef_g", "h", ""), "_", n_max=3, tokens_only=TRUE, omit_empty=TRUE))
```

* **(NEW FEATURE)** [#107](https://github.com/Rexamine/stringi/issues/107):
  `stri_split_*` now allow setting an `omit_empty=NA` argument.

```{r}
stri_split_fixed("a_b_c__d", "_", omit_empty=FALSE)
stri_split_fixed("a_b_c__d", "_", omit_empty=TRUE)
stri_split_fixed("a_b_c__d", "_", omit_empty=NA)
```

* **(NEW FEATURE)** [#106](https://github.com/Rexamine/stringi/issues/106):
  `stri_split` and `stri_extract_all` gained a `simplify`
  argument (if `TRUE`, then `stri_list2matrix(..., byrow=TRUE)`
  is called on the resulting list.
  
```{r}
stri_split_fixed(c("ab,c", "d,ef,g", ",h", ""), ",", omit_empty=TRUE, simplify=TRUE)
stri_split_fixed(c("ab,c", "d,ef,g", ",h", ""), ",", omit_empty=FALSE, simplify=TRUE)
stri_split_fixed(c("ab,c", "d,ef,g", ",h", ""), ",", omit_empty=NA, simplify=TRUE)
```

* **(NEW FUNCTION)** [#77](https://github.com/Rexamine/stringi/issues/77):
  `stri_rand_lipsum` generates
  (pseudo)random dummy [*lorem ipsum*](http://lipsum.com/) text.
  
```{r}
cat(sapply(
   stri_wrap(stri_rand_lipsum(3), 80, simplify=FALSE),
   stri_flatten, collapse="\n"), sep="\n\n")
```

* **(NEW FEATURE)** [#98](https://github.com/Rexamine/stringi/issues/98):
  `stri_trans_totitle` gained a `opts_brkiter`
  parameter; it indicates which ICU BreakIterator should be used when
  performing case mapping.
  
```{r}
stri_trans_totitle("GOOD-OLD cOOkiE mOnSTeR IS watCHinG You. Here HE comes!",
    stri_opts_brkiter(type="word")) # default boundary
stri_trans_totitle("GOOD-OLD cOOkiE mOnSTeR IS watCHinG You. Here HE comes!",
    stri_opts_brkiter(type="sentence"))
```

* **(NEW FEATURE)** `stri_wrap` gained a new parameter: `normalize`.

* **(BUGFIX)** [#86](https://github.com/Rexamine/stringi/issues/86):
  `stri_*_fixed`, `stri_*_coll`, and `stri_*_regex` could
  give incorrect results if one of search strings were of length 0.

* **(BUGFIX)** [#99](https://github.com/Rexamine/stringi/issues/99):
  `stri_replace_all` did not use the `replacement` arg.

* **(BUGFIX)** [#94](https://github.com/Rexamine/stringi/issues/94):
  `R CMD check` should no longer fail if `icudt` download failed.

* **(BUGFIX)** [#112](https://github.com/Rexamine/stringi/issues/112):
  Some of the objects were not PROTECTed from
  being garbage collected, which might have caused spontaneous SEGFAULTS.

* **(BUGFIX)** Some collator's options were not passed correctly to ICU services.

* **(BUGFIX)** Memory leaks causes as detected by
  `valgrind --tool=memcheck --leak-check=full` have been removed.

* **(DOCUMENTATION)** Significant extensions/clean ups in the [`stringi`](http://stringi.rexamine.com) manual.
   
    [Check out](http://docs.rexamine.com/R-man/stringi/stringi-package.html)
   yourself. In particular, take a glimpse at
   [`stringi-search-regex`](http://docs.rexamine.com/R-man/stringi/stringi-search-regex.html),
   [`stringi-search-charclass`](http://docs.rexamine.com/R-man/stringi/stringi-search.html)
   and, more generally, at
   [`stringi-search`](http://docs.rexamine.com/R-man/stringi/stringi-search-charclass.html).
 
 
Enjoy! Any comments and suggestions are [welcome](https://github.com/Rexamine/stringi/issues).