(String/text processing)++: stringi 0.2-3 released
==================================================

A new release of the [`stringi`](http://stringi.rexamine.com) package is
available on [CRAN](http://cran.r-project.org/web/packages/stringi/index.html)
(please wait a few days for Windows and OS X binary builds).

**`stringi` is a package providing** (but definitely not limiting to) equivalents
of nearly all the **character string processing functions** known
from base R. While developing the package we had
**high performance** and **portability** of its facilities in our minds.

`stringi`'s user interface is **inspired by
and consistent with** that of Hadley Wickham's great
**[`stringr`](http://cran.r-project.org/web/packages/stringr/index.html) package**.
Quoting its [README](http://cran.r-project.org/web/packages/stringr/README.html),
`stringr` (and "hence" `stringi`):

> * Processes factors and characters in the same way,
> * Gives functions consistent names and arguments,
> * Simplifies string operations by eliminating options that you don't need 95% of the time,
> * Produces outputs than can easily be used as inputs. This includes ensuring that missing inputs result in missing outputs, and zero length inputs result in zero length outputs,
> * Completes R's string handling functions with useful functions from other programming languages.

### Some problems with base R functions ###

While base R as well as `stringr` functions are great for simple
text processing tasks, dealing with more complex ones
(such as natural language processing) may be a bit problematic.

First of all, some time ago we mentioned in
[our blog post](http://www.rexamine.com/2013/04/properly-internationalized-regular-expressions-in-r/)
that regex search may provide **different outputs on different platforms**.
For example, Polish letters such as ą, ę, ś etc. are correctly
captured with `[[:alpha:]]` by the (default) ERE engine on Linux (native encoding=UTF-8),
while on Windows the results are quite surprising.
(A year ago my students got (of course, initially)
very bad marks from a Polish text processing
task just because they had written their R scripts on Windows while I
ran them on Linux.) :-)

Secondly, natural language processing relies on **a set of very complex,
locale-specific rules**. However, the rules available
(via e.g. [glibc](http://www.gnu.org/software/libc/manual/html_node/Character-Handling.html))
in base R string functions may sometimes give incorrect results.
For example, when we convert German `ß` (es-zett/double small s)
character to upper case, we rather expect ``SS`` in result than:

```{r,width=40}
toupper("groß") # GROSS? No...
```

Moreover, let's assume that we are asked to sort a character
vector according to the rules specific to the Slovak language.
Here, quite interestingly, the word `hladný` (hungry)
can be found in a dictionary *before* the word `chladný` (cold).
Of course, as not everyone works in a Slovak locale,
we don't expect to obtain a proper order immediately:

```{r}
sort(c("hladný", "chladný"))
```

In order to obtain a proper order, we should temporarily
switch to a Slovak "environment":

```{r,results=3}
oldlocale <- Sys.getlocale("LC_COLLATE")
Sys.setlocale("LC_COLLATE", "sk_SK")
sort(c("hladný", "chladný"))
Sys.setlocale("LC_COLLATE", oldlocale)
```

This code works on my Linux, but is **not portable**. It's because:

1. Other Linux users may not have Slovak rule-base installed (and
not everyone has abilities to do it on his/her own).
2. Windows users don't use BCP 47-based locale names. There,
LCID `Slovak_Slovakia.1250` is appropriate.

And so on.

### `stringi` facilities ###

In order to overcome such problems we decided to reimplement
each string processing function from scratch
(of course, purely in C++). The internationalization and globalization
support, as well as many string processing facilities
(like regex searching) is guaranteed by the well-known and established
**IBM's [ICU4C](http://site.icu-project.org/) library** (refer to
ICU's website for more details).

Here is a very general list of the **most important
features** available in the current version of `stringi`:

* string searching:
    * with ICU (Java-like) regular expressions,
    * ICU USearch-based locale-aware string searching
      (quite slow, but working properly e.g. for non-Unicode normalized strings),
    * very fast, locale-independent byte-wise pattern matching;
* joining and duplicating strings;
* extracting and replacing substrings;
* string trimming, padding, and text wrapping (e.g. with Knuth's dynamic word wrap algorithm);
* text transliteration;
* text collation (comparing, sorting);
* text boundary analysis (e.g. for extracting individual words);
* random string generation;
* Unicode normalization;
* character encoding conversion and detection;

and many more.

### Showcase ###

Here's a bunch of examples.

```{r}
library(stringi)
```

* Proper `NA` handling:

```{r,message=FALSE}
stri_length(c("aaa", NA, ""))
```

* "Deep" vectorization:

```{r}
stri_replace_all_fixed(c("aba", "bab"), c("a", "b"), c("c", "d")) # 1-1-1 and 2-2-2
stri_replace_all_fixed(c("aba", "bab"), "a", "c") # 1-1-1 and 2-1-1
stri_replace_all_fixed("aba", c("a", "b"), "c") # 1-1-1 and 1-2-1
stri_replace_all_fixed("aba", "a", c("c", "d")) # 1-1-1 and 1-1-2
```

(all the functions are vectorized w.r.t most of their arguments)

* Easy-to-use, portable locale selection:

```{r}
stri_sort(c("hladný", "chladný"), opts=stri_opts_collator(locale="sk_SK"))
```

* Proper transliteration rules:

```{r}
stri_trans_toupper("Groß")
```


**In our upcoming blog posts we will present some exciting
features of `stringi`.**
They are definitely worth to be discussed separately! Stay tuned.


### Performance ###

And some benchmarks.

* String sorting:

```{r,bench_sort,cache=TRUE}
set.seed(123L)
library(microbenchmark)
x <- stri_rand_strings(100000, 10) # 10000 random ASCII "words" of length 10 each
head(x, 5)
microbenchmark(sort(x), stri_sort(x))
```

* String joining:

```{r,bench_paste,cache=TRUE}
microbenchmark(paste(x, collapse=", "), stri_paste(x, collapse=", "))
```

* Searching for a fixed pattern:

```{r,bench_search,cache=TRUE}
set.seed(123L)
y <- stri_rand_strings(10000, 10, "[ACGT]") # 10000 random "genomes" of length 10
head(y, 5)
microbenchmark(
   grepl("ACCA", y),
   grepl("ACCA", y, fixed=TRUE),
   grepl("ACCA", y, perl=TRUE),
   stri_detect_fixed(y, "ACCA"),
   stri_detect_regex(y, "ACCA")
)
```

* Determining a substring:

```{r,bench_unique,cache=TRUE}
microbenchmark(substr(y, 2, 4), stri_sub(y, 2, 4))
```

As a **rule of thumb**: `stringi` functions should often be
faster than the R ones for long ASCII and UTF-8 strings.
They often have poorer performance for short 8-bit encoded ones.


### More information ###

For **more information** check out the `stringi`
package [website](http://www.rexamine.com/resources/stringi/)
and its [on-line documentation](http://docs.rexamine.com/R-man/stringi/stringi-package.html).

For **bug reports and feature requests** visit our [GitHub profile](https://github.com/Rexamine/stringi/).

In the future versions of `stringi` we plan to include:

* rule-based number formatting (number spell-out, e.g. `123 -> one hundred twenty three`);
* date and time formatting/parsing;
* access to the Unicode Character database;
* functions to read/write text files (with automatic encoding detection);
* and many more.

**Any comments and suggestions are warmly welcome.**

Have fun!

--
[Marek Gagolewski](http://gagolewski.rexamine.com)

### Change-log ###


**Notable changes** since the previous CRAN release (v0.1-25):

* [IMPORTANT CHANGE] `stri_cmp*` now do not allow for passing `opts_collator=NA`.
From now on, `stri_cmp_eq`, `stri_cmp_neq`, and the new operators
`%===%`, `%!==%`, `%stri===%`, and `%stri!==%` are locale-independent operations,
which base on code point comparisons. New functions `stri_cmp_equiv`
and `stri_cmp_nequiv` (and from now on also `%==%`, `%!=%`, `%stri==%`,
and `%stri!=%`) test for canonical equivalence.

* [IMPORTANT CHANGE] `stri_*_fixed` search functions now perform
a locale-independent exact (bytewise, of course after conversion to UTF-8)
pattern search. All the Collator-based, locale-dependent search routines
are now available via `stri_*_coll`. The reason for this is that
ICU USearch has currently very poor performance and in many search tasks
in fact it is sufficient to do exact pattern matching.

* [IMPORTANT CHANGE] `stri_enc_nf*` and `stri_enc_isnf* `function families
have been renamed to `stri_trans_nf*` and `stri_trans_isnf*`, respectively.
This is because they deal with text transforming, and not with character
encoding. Moreover, all such operation may be performed by
ICU's Transliterator (see below).

* [IMPORTANT CHANGE] `stri_*_charclass` search functions now
rely solely on ICU's UnicodeSet patterns. All previously accepted
charclass identifiers became invalid. However, new patterns
should now be more familiar to the users (they are regex-like).
Moreover, we observe a very nice performance gain.

* [IMPORTANT CHANGE] stri_sort now does not include NAs
in output vectors by default, for compatibility with `sort()`.
Moreover, currently none of the input vector's attributes are preserved.

* [NEW FUNCTION] `stri_trans_general`, `stri_trans_list` gives access
to ICU's Transliterator: may be used to perform very general
text transforms.

* [NEW FUNCTION `stri_split_boundaries` utilizes ICU's BreakIterator
to split strings at specific text boundaries. Moreover,
`stri_locate_boundaries` indicates positions of these boundaries.

* [NEW FUNCTION] `stri_extract_words` uses ICU's BreakIterator to
extract all words from a text. Additionally, `stri_locate_words`
locates start and end positions of words in a text.

* [NEW FUNCTION] `stri_pad`, `stri_pad_left`, `stri_pad_right`, `stri_pad_both`
pads a string with a specific code point.

* [NEW FUNCTION] `stri_wrap` breaks paragraphs of text into lines.
Two algorihms (greedy and minimal-raggedness) are available.

* [NEW FUNCTION] `stri_unique` extracts unique elements from
a character vector.

* [NEW FUNCTIONS] `stri_duplicated` any `stri_duplicated_any`
determine duplicate elements in a character vector.

* [NEW FUNCTION] `stri_replace_na` replaces `NA`s in a character vector
with a given string, useful for emulating e.g. R's `paste()` behavior.

* [NEW FUNCTION] `stri_rand_shuffle` generates a random permutation
of code points in a string.

* [NEW FUNCTION] `stri_rand_strings` generates random strings.

* [NEW FUNCTIONS] New functions and binary operators for string comparison:
`stri_cmp_eq`, `stri_cmp_neq`, `stri_cmp_lt`, `stri_cmp_le`, `stri_cmp_gt`,
`stri_cmp_ge`, `%==%`, `%!=%`, `%<%`, `%<=%`, `%>%`, `%>=%`.

* [NEW FUNCTION] `stri_enc_mark` reads declared encodings of character strings
as seen by `stringi`.

* [NEW FUNCTION] `stri_enc_tonative(str)` is an alias to
`stri_encode(str, NULL, NULL)`.

* [NEW FEATURE] `stri_order` and `stri_sort` now have an additional argument
`na_last` (defaults to `TRUE` and `NA`, respectively).

* [NEW FEATURE] `stri_replace_all_charclass` now has `merge` arg
(defaults to `FALSE` for backward-compatibility). It may be used
to e.g. replace sequences of white spaces with a single space.

* [NEW FEATURE] `stri_enc_toutf8` now has a new `validate` arg (defaults
to `FALSE` for backward-compatibility). It may be used in a (rare) case
in which a user wants to fix an invalid UTF-8 byte sequence.
stri_length (among others) now detect invalid UTF-8 byte sequences.

* [NEW FEATURE] All binary operators `%???%` now also have aliases `%stri???%`.

* `stri_*_fixed` now use a tweaked Knuth-Morris-Pratt search algorithm,
which improves the search performance drastically.

* Significant performance improvements in `stri_join`, `stri_flatten`,
`stri_cmp`, `stri_trans_to*`, and others.


Refer to [NEWS](https://raw.githubusercontent.com/Rexamine/stringi/master/NEWS)
for a complete list of changes, new features and bug fixes.
