Transforms/Transliteration
========================================================

Transforms provide a general-purpose package for processing Unicode text. They are a powerful and flexible mechanism for handling a variety of different tasks, including:

    Uppercase, Lowercase, Titlecase, Full/Halfwidth conversions
    (Full (Language-Specific) Case Mapping)
    
case mapping can change the number of code points and/or code units of a string,
is language-sensitive (results may differ depending on language), and
is context-sensitive (a character in the input string may map differently depending on surrounding characters).



    Normalization
Normalization is used to convert text to a unique, equivalent form
Normalization allows for easier sorting and searching of text


```{r}
library(stringi)
stri_trans_nfd(stri_unescape_unicode("\\u0105")) # Polish a with ogonek -> a, ogonek
stri_trans_nfkc(stri_unescape_unicode("\\ufdfa")) # 1 codepoint -> 18 codepoints
```

    Hex and Character Name conversions

    Script to Script conversion
    
    Script Transliteration is the general process of converting characters from one script to another. 
    it is not translation
    
http://userguide.icu-project.org/transforms/general

 creates transforms, restricts their effective range, and combines them together.
 
 
```{r}
library(stringi)
stri_trans_list()
stri_trans_general("w szczebrzeszynie chrząszcz brzmi w trzcinie", "pl-pl_fonipa")
stri_trans_general("w szczebrzeszynie chrząszcz brzmi w trzcinie", "pl-pl_fonipa; ipa-xsampa") # XSampa ASCII-range representation of IPA characters.
stri_trans_general("zażółć gęślą jaźń", "[^\\u0000-\\u007f] any-hex") # filtered
stri_trans_general("zażółć gęślą jaźń", "[^\\u0000-\\u007f] any-hex/xml")
stri_trans_general("zażółć gęślą jaźń", "latin-ascii")
stri_trans_general("ą1©,", "any-name")
stri_trans_general("\\N{LATIN SMALL LETTER SHARP S}", "name-any")
stri_trans_general("Let's go -- she said", "any-publishing")
stri_trans_general("zażółć gęślą jaźń", "[\\p{Z}] remove")
stri_trans_general("zażółć gęślą jaźń", "[\\p{Z}] remove")
stri_trans_general("stringi", "latin-greek") # script transliteration
stri_trans_general("Пётр Ильич Чайковский", "cyrillic-latin") # script transliteration
stri_trans_general("Пётр Ильич Чайковский", "cyrillic-latin; nfd; [:nonspacing mark:] remove; nfc") # and remove accents
```
