(Sec:io)=
Input and Output
================


```{r echo=FALSE, results="asis", message=FALSE}
source("common.R")
opts_chunk$set(cache.path="cache/input_output_")
```


This section deals with some more advanced topics related to the
operability of text processing applications between different platforms.
In particular, we discuss how to assure that data read from various
input connections are interpreted in the correct manner.

(Sec:codepoints)=
Dealing with Unicode Code Points
--------------------------------

The Unicode Standard (as well as the Universal Coded Character Set,
i.e., ISO/IEC 10646) currently defines over 140,000 abstract characters
together with their corresponding code points -- integers between 0 and
1,114,111 (or 0000 and 10FFFF in hexadecimal notation,
see <https://www.unicode.org/charts/>). In particular, here is the
number of the code points in some popular categories (compare
{ref}`Sec:RegexIndividualChars`), such as letters, numbers, and
the like.

```{r }
z <- c("\\p{L}", "\\p{Ll}", "\\p{Lu}", "\\p{N}", "\\p{P}", "\\p{S}",
  "\\w", "\\d", "\\s")
structure(stri_count_regex(stri_enc_fromutf32(
  setdiff(1:0x10ffff, c(0xd800:0xf8ff))), z), names=z)
```

Yet, most of the code points are still unallocated. The Unicode standard
is occasionally updated, e.g., the most recent versions were
supplemented with over 1,000 emojis.

The first 255 code points are identical to the ones defined by ISO/IEC
8859-1 (ISO Latin-1; "Western European"), which itself extends US-ASCII
(codes ≤ U+007f). For instance, the code point that we
are used to denoting as U+007A (the "U+" prefix is followed by a
sequence of hexadecimal digits; U+007A corresponds to decimal 122)
encodes the lower case letter "z". To input such a code point in R, we
write:

```{r }
"\u007A"  # or "\U0000007A"
```

For communicating with *ICU* and other libraries, we may need to escape
a given string, for example, as follows (recall that to input a
backslash in R, we must escape it with another backslash).

```{r }
x <- "zß你好"
stri_escape_unicode(x)
```

Even though some output devices might
be unable to display certain code points correctly (due to, e.g.,
missing fonts), the correctness of their processing with *stringi* is
still guaranteed by *ICU*.

(Sec:encoding)=
Character Encodings
-------------------

When storing strings in RAM or on the disk, we need to decide upon the
actual way of representing the code points as sequences of bytes. The
two most popular encodings in the Unicode family are UTF-8 and UTF-16:

```{r }
x <- "abz0ąß你好!"
stri_encode(x, to="UTF-8", to_raw=TRUE)[[1]]
stri_encode(x, to="UTF-16LE", to_raw=TRUE)[[1]]  # little-endian
```

R's current platform-default encoding, which we'll refer to as the
native encoding, is defined via the `LC_CTYPE` locale category in
`Sys.getlocale()`. This is the representation assumed, e.g., when
reading data from the standard input or from files (e.g., when `scan()`
is called). For instance, Central European versions of Windows will
assume the "`windows-1250`" code page. MacOS as well as most Linux boxes
work with UTF-8 by default[^footucrt].

All strings in R have an associated encoding mark which can be read by
calling `Encoding()` or, more conveniently, `stri_enc_mark()`. Most
importantly, strings in ASCII, ISO-8859-1 ("`latin1`"), UTF-8, and the
native encoding can coexist. Whenever a non-Unicode string is passed to
a function in *stringi*, it is silently converted to UTF-8 or UTF-16,
depending on the requested operation (some *ICU* services are only
available for UTF-16 data). Over the years, this has proven a robust,
efficient, and maximally portable design choice -- Unicode can be
thought of as a superset of every other encoding. Moreover, in order to
guarantee the correctness and high performance of the string processing
pipelines, *stringi* always[^footuftout] outputs UTF-8 data.

[^footucrt]: It is expected that future R releases will support UTF-8 natively
    thanks to the Universal C Runtime (UCRT) that is available for
    Windows 10.

[^footuftout]: Okay, with a few obvious exceptions, such as `stri_encode()`.



(Sec:read_lines)=
Reading and Writing Text Files
------------------------------

According to a report by
[W3Techs](https://w3techs.com/technologies/cross/character_encoding/ranking),
as of 2021-09-28, 97.3% of
websites use UTF-8. Nevertheless, other encodings can still be
encountered.



If we know the encoding of a text file in advance, `stri_read_lines()`
can be used to read the data in a manner similar to the built-in
`readLines()` function (but with a much easier access to encoding
conversion):

For instance, below we read a text file
(see <https://github.com/gagolews/stringi/tree/master/docs/_static/vignette>)
encoded in ISO-8859-1:

```{r }
x <- stri_read_lines("ES_latin1.txt", encoding="ISO-8859-1")
head(x, 4)  # x is in UTF-8 now
```

We can call `stri_write_lines()` to write the contents of a character
vector to a file (each string will constitute a separate text line),
with any output encoding.


Detecting Encodings
-------------------

If a file's encoding is not known in advance, there are a
certain functions that can aid in encoding detection. First, we can read
the resource in form of a raw-type vector:

```{r }
x <- stri_read_raw("ES_latin1.txt")
head(x, 24)  # vector of type raw
```

Then, to guess the encoding, we can call, e.g.:

```{r }
stri_enc_isascii(x)
stri_enc_isutf8(x)   # false positives are possible
```

Alternatively, we can use:

```{r }
stri_enc_detect(x)  # based on heuristics
```

Nevertheless, encoding detection is an operation that relies on
heuristics, therefore there is a chance that the output might be
imprecise or even misleading.


Converting Between Encodings
----------------------------


Knowing the desired source and destination encoding precisely,
`stri_encode()` can be called to perform the conversion. Contrary to the
built-in `iconv()`, which relies on different underlying libraries, the
current function is portable across operating systems.

```{r }
y <- stri_encode(x, from="ISO-8859-1", to="UTF-8")
```

`stri_enc_list()` provides a list of supported encodings and their
aliases in many different forms. Encoding specifiers are normalised
automatically, e.g., `"utf8"` is a synonym for `"UTF-8"`.

Splitting the output into text lines gives:

```{r }
tail(stri_split_lines1(y), 4)  # spoiler alert!
```

(Sec:normalisation)=
Normalising Strings
-------------------

In the section on {ref}`Sec:Equivalence` we've provided some examples of
canonically equivalent strings whose code point representation was
different. Unicode [normalisation](https://www.unicode.org/reports/tr15/)
forms C (Canonical composition, NFC)
and D (Canonical decomposition, NFD) can be applied so that they'll
compare equal using bytewise matching.

```{r }
x <- "a\u0328 ą"   # a, combining ogonek, space, a with ogonek
stri_enc_toutf32(  # code points as decimals
  c(x, stri_trans_nfc(x), stri_trans_nfd(x)))
```

Above we see some example code points before, after NFC, and after NFD
normalisation, respectively.

It might be a good idea to always normalise all the strings read from
external sources (files, URLs) with NFC.

Compatibility composition and decomposition normalisation forms (NFKC
and NFKD, respectively) are also available if the removal of the
formatting distinctions (font variants, subscripts, superscripts, etc.)
is desired. For example:

```{r }
stri_trans_nfkd("r²︷")
```

Such text might be much easier to process or analyse statistically.
