---
title: "stringi 0.4-1 released - fast, portable, consistent character string processing"
author: "Marek Gagolewski"
date: "14.12.2014"
output: html_document
---

A new release of the [`stringi`](https://github.com/Rexamine/stringi/) package is
available on [CRAN](http://cran.r-project.org/web/packages/stringi/index.html)
(please wait a few days for Windows and OS X binary builds).

```{r}
# install.packages("stringi") or update.packages()
library("stringi")
```

Here's a list of changes in version 0.4-1. In the current release,
we particularly focused on making the package's interface
more consistent with that of the well-known
[`stringr`](https://github.com/hadley/stringr) package.
For a general overview of `stringi`'s facilities and base R string processing
issues, see e.g. [here](http://www.rexamine.com/2014/05/stringi-0-2-3-released/).


* **(IMPORTANT CHANGE)** `n_max` argument in `stri_split_*()` has been renamed `n`.

* **(IMPORTANT CHANGE)** `simplify=FALSE` in `stri_extract_all_*()` and
  `stri_split_*()` now calls `stri_list2matrix()` with `fill=""`.
  `fill=NA_character_` may be obtained by using `simplify=NA`.

* **(IMPORTANT CHANGE, NEW FUNCTIONS)** [#120](https://github.com/Rexamine/stringi/issues/120): `stri_extract_words` has been
  renamed `stri_extract_all_words` and `stri_locate_boundaries` -
  `stri_locate_all_boundaries` as well as `stri_locate_words` -
  `stri_locate_all_words`. New functions are now available:
  `stri_locate_first_boundaries`, `stri_locate_last_boundaries`,
  `stri_locate_first_words`, `stri_locate_last_words`,
  `stri_extract_first_words`, `stri_extract_last_words`.
  
```{r}
# uses ICU's locale-dependent word break iterator
stri_extract_all_words("stringi: THE string processing package for R")
```

* **(IMPORTANT CHANGE)** [#111](https://github.com/Rexamine/stringi/issues/111): `opts_regex`, `opts_collator`, `opts_fixed`, and
  `opts_brkiter` can now be supplied individually via `...`.
  In other words, you may now simply call e.g.
  
```{r}
stri_detect_regex(c("stringi", "STRINGI"), "stringi", case_insensitive=TRUE)
```

instead of:

```{r}
stri_detect_regex(c("stringi", "STRINGI"), "stringi", opts_regex=stri_opts_regex(case_insensitive=TRUE))
```


* **(NEW FEATURE)** [#110](https://github.com/Rexamine/stringi/issues/110): Fixed pattern search engine's settings can
  now be supplied via `opts_fixed` argument in `stri_*_fixed()`,
  see `stri_opts_fixed()`. A simple (not suitable for natural language
  processing) yet very fast `case_insensitive` pattern matching can be
  performed now. `stri_extract_*_fixed` is again available.

* **(NEW FEATURE)** [#23](https://github.com/Rexamine/stringi/issues/23): `stri_extract_all_fixed`, `stri_count`, and
  `stri_locate_all_fixed` may now also look for overlapping pattern
  matches, see `?stri_opts_fixed`.
  
```{r}
stri_extract_all_fixed("abaBAaba", "ABA", case_insensitive=TRUE, overlap=TRUE)
```

* **(NEW FEATURE)** [#129](https://github.com/Rexamine/stringi/issues/129): `stri_match_*_regex` gained a `cg_missing` argument.

* **(NEW FEATURE)** [#117](https://github.com/Rexamine/stringi/issues/117): `stri_extract_all_*()`, `stri_locate_all_*()`,
  `stri_match_all_*()` gained a new argument: `omit_no_match`.
  Setting it to `TRUE` makes these functions compatible with their
  `stringr` equivalents.

* **(NEW FEATURE)** [#118](https://github.com/Rexamine/stringi/issues/118): `stri_wrap()` gained `indent`, `exdent`, `initial`,
  and `prefix` arguments. Moreover Knuth's dynamic word wrapping algorithm
  now assumes that the cost of printing the last line is zero, see [#128](https://github.com/Rexamine/stringi/issues/128).
  
```{r}
cat(stri_wrap(stri_rand_lipsum(1), 40, 2.0), sep="\n")
```

* **(NEW FEATURE)** [#122](https://github.com/Rexamine/stringi/issues/122): `stri_subset()` gained an `omit_na` argument.

```{r}
stri_subset_fixed(c("abc", NA, "def"), "a")
stri_subset_fixed(c("abc", NA, "def"), "a", omit_na=TRUE)
```

* **(NEW FEATURE)** `stri_list2matrix()` gained an `n_min` argument.

* **(NEW FEATURE)** [#126](https://github.com/Rexamine/stringi/issues/126): `stri_split()` now is also able to act
  just like `stringr::str_split_fixed()`.
  
```{r}
stri_split_regex(c("bab", "babab"), "a", n = 3, simplify=TRUE)
```

* **(NEW FEATURE)** [#119](https://github.com/Rexamine/stringi/issues/119):  `stri_split_boundaries()` now have
  `n`, `tokens_only`, and `simplify` arguments. Additionally,
  `stri_extract_all_words()` is now equipped with `simplify` arg.
  
* **(NEW FEATURE)** [#116](https://github.com/Rexamine/stringi/issues/116): `stri_paste()` gained a new argument:
  `ignore_null`. Setting it to `TRUE` makes this function more compatible
  with `paste()`.
  
```{r}
for (test in c(TRUE, FALSE))
   print(stri_paste("a", if (test) 1:9, ignore_null=TRUE))
```
 
Enjoy! Any comments and suggestions are [welcome](https://github.com/Rexamine/stringi/issues).
